The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-11-05 submission/manuscript.Rmd
shigeki abstract edits
discussion small edit

The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice aberrantly express many tissue-specific genes within the brain, the majority of which are unique to the germline. By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature _Kdm5c_-KO brain. KDM5C binds germline gene promoter CGIs to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs). However, the majority of late-stage spermatogenesis genes expressed within the _Kdm5c_-KO brain did not harbor promoter CGIs. These CGI-free germline genes were not bound by KDM5C and instead expressed through ectopic activation by RFX transcription factors. Furthermore, KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.

Altogether, these data revealed distinct regulatory classes of germline genes and sex-biased silencing mechanisms in somatic cells.

@@

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.

Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes employ the same silencing mechanisms, such as those involved in the later stages of oogenesis and spermatogenesis. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e. sex and tissue environment) influences the manifestation of germline gene misexpression. Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.

@@

To illuminate KDM5C's role in tissue identity, we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, enabling genome-wide analysis of germline gene silencing mechanisms for the first time. Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through CGI DNA methylation. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled distinct mechansisms governing the misexpression of distinct germline gene classes in somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

>>>>

The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that drive germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGIs). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice aberrantly express many tissue-specific genes within the brain, the majority of which are unique to the germline. By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature brain. KDM5C binds CGIs within germline gene promoters to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs). However, the majority of late-stage spermatogenesis genes expressed within the _Kdm5c_-KO brain did not harbor promoter CGIs. These CGI-free germline genes were not bound by KDM5C and instead expressed through ectopic activation by RFX transcription factors. Furthermore, germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing. Altogether, these data revealed distinct regulatory classes of germline genes and sex-biased silencing mechanisms in somatic cells.

@@

The separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Germline gene promoters initially gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes employ the same silencing mechanisms, such as those involved in the later stages of oogenesis and spermatogenesis. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e. sex and tissue environment) influences the manifestation of germline gene misexpression. Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.

@@

To illuminate KDM5C's role in tissue identity, we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, enabling genome-wide analysis of germline gene silencing mechanisms for the first time. We identified two classes of germline genes based on their promoter CpG island content, which are dysregulated with KDM5C loss by distinct mechanisms and in a sex-biased manner.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through CGI DNA methylation. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled divergent mechanisms governing the misexpression of distinct germline gene classes in somatic lineages.
